QDNAseq segmentation

### 2019-10-25 QDNAseq segmentation.Rmd
changing cv to select different alpha values

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)

@@

training.raw = read_tsv('~/Data/BarrettsProgressionRisk/Analysis/multipcf_perPatient/pre_seg_medians.tsv', col_types = 'cdd') %>% dplyr::rename(sd='SD',sample='Sample') %>% mutate(cohort = 'train')

v15.file= '~/Data/BarrettsProgressionRisk/Analysis/validation/pre_seg_disp.tsv'
all.val15 = read_tsv(v15.file, col_types = 'ddc') %>% dplyr::select(sample,median,sd) %>% mutate(cohort='val')
```

@@

I'm looking at the variance in the raw data immediately following QDNAseq processing. The training data was all run with a 15kb bin size so in the initial runs of the validation data all samples were run with that bin size.

## Comparing the distributions

Training vs Validation set

```{r, echo=F}
ggplot(bind_rows(training.raw, all.val15)) + xlim(0.95,1.05) + ylim(0,1.0) + 
  geom_point(aes(median,sd,color=cohort)) + 
  theme_minimal() + labs(title='Median vs SD for available samples')
```

The median is pretty close to the same, but there is a larger spread in the validation dataset.

```{r}
tv15 = bind_rows(training.raw, all.val15) %>% mutate(cohort = factor(cohort))

ggplot(tv15, aes(median)) + facet_grid(~cohort, scales = 'free_x') +# xlim(0.95,1.05) +
  geom_histogram(bins=20, color='lightblue')

ggplot(tv15, aes(cohort, median, group=cohort, fill=cohort)) + ylim(0.95,1.05) +
  geom_jitter(width=0.2) + geom_boxplot()

wilcox.test(all.val15$median, training.raw$median)
```

The difference is clearer in the std. deviation of the two cohorts, even when the extreme outliers are removed.

```{r}
ggplot(tv15, aes(sd)) + facet_grid(~cohort) + xlim(0,1) +
  geom_histogram(bins=50, color='lightblue')

ggplot(tv15, aes(cohort, sd, group=cohort, fill=cohort)) + ylim(0,1) + 
  geom_jitter(width=0.2) + geom_boxplot(outlier.colour = NA)

wilcox.test(all.val15$sd, training.raw$sd)
```

### Per sample the shift is event more obvious.

Dotted line is median(sd).
```{r, echo=F}
ggplot(tv15, aes(x=1:nrow(tv15),y=sd)) + ylim(0,1.0) + 
  geom_point(aes(color=cohort)) + geom_hline(yintercept = median(training.raw$sd), linetype='dashed') +
  #geom_hline(yintercept = median(training.raw$sd)+mad(all.val15$sd), color='grey39') +
  theme(axis.text.x = element_blank()) +
  theme_minimal() + labs(title='Training and validation cohorts (15kb bins)')
```

## Altering the QDNAseq bin size

Looking at different bin sizes in the validation cohort. Ellie's bam files are not as easily recovered so I'm using the training cohort as the standard and altering the validation cohort. The median variance decreases as I increase the bin size.
```{r}
files = grep('ellie',list.files('~/Data/BarrettsProgressionRisk/data', 'prepped_dist.tsv', recursive = T, full.names = T),invert = T, value=T)
val = do.call(bind_rows, lapply(files, function(f) {
  read_tsv(f, col_types = 'cdd') %>% mutate('seg.size'=basename(dirname(f)))
})) %>% mutate(seg.size = factor(seg.size)) %>% arrange(sample, seg.size)

val = val %>% mutate(kb = as.integer(sub('kb','',seg.size)), seg.size = factor(seg.size, levels=c('15kb','30kb','50kb','100kb','500kb'), ordered=T)) %>% arrange(kb) %>% mutate(index = row_number())

samples = val %>% group_by(sample) %>% tally %>% dplyr::filter(n == 4) %>% select(sample) %>% pull

#val = val %>% group_by(sample) %>% mutate(diff = dist(sd))
ggplot(val,  aes(index, sd, color=seg.size)) + ylim(0,1) + 
  geom_point() + 
  theme(axis.text.x=element_blank()) + theme_minimal()

ggplot(val %>% arrange(kb),  aes(seg.size, sd, color=seg.size)) + ylim(0,1) + 
  geom_jitter(width=0.2) + geom_boxplot(alpha=0.8, outlier.colour = NA) +
  theme_minimal() + theme(axis.text.x=element_blank(), legend.position = 'none')

val = val %>% filter(kb > 15)

tv15 = tv15 %>% mutate(seg.size = '15kb', kb=15) %>% dplyr::select(sample,cohort,seg.size,kb,sd,median)
val = val %>% mutate(cohort = paste('val',seg.size)) %>% dplyr::select(sample,cohort,seg.size,kb,sd,median)

merged = bind_rows(val,tv15)

merged = merged %>% mutate(
  seg.size = factor(seg.size, levels=c('train','15kb','30kb','50kb','100kb','500kb'), ordered=T),
  cohort = factor(cohort, levels=c('train','val','val 30kb','val 50kb','val 100kb','val 500kb'), ordered=T)
  ) %>% arrange(seg.size)

ggplot(merged, aes(seg.size, sd, group=seg.size, fill=seg.size)) + facet_grid(~cohort, scales='free_x') + ylim(0,1) +
  geom_jitter(width=0.2) + geom_boxplot(outlier.colour = NA) + 
  scale_fill_brewer(palette = 'Set1') + theme_minimal()

@@

So it appears that I can decrease the variance by increasing the bin size.  But it doesn't necessarily make sense to do it across the board.  I can take the median of the training data and call that the cutoff for deciding.

```{r}
ggplot(merged, aes(x=1:nrow(merged),y=sd)) + ylim(0,1.0) + 
  geom_point(aes(color=cohort)) + geom_hline(yintercept = median(training.raw$sd), linetype='dashed') +
  geom_hline(yintercept = median(training.raw$sd)+0.5*mad(all.val15$sd), color='grey39') +
  theme(axis.text.x = element_blank()) +
  theme_minimal()

out.samples = all.val15 %>% filter(sd > median(training.raw$sd))# +0.5*mad(all.val15$sd)))

In the validation dataset, starting at 15kb `r nrow(out.samples)`:`r nrow(all.val15)` samples have a variance greater than the median.

```{r, echo=F}
preds15 = do.call(bind_rows, lapply(list.files('~/Data/BarrettsProgressionRisk/Analysis/validation/multipcf/predictions_model_5e6/0.9','predictions',full.names=T, recursive=T), function(f) read_tsv(f,col_types='cddcccDcccc')))

#preds15 %>% filter(Sample %in% out.samples$sample) %>% group_by(Risk) %>% tally
```

## Re-process dynamically

I re-processed all of the validation data with this in mind.  I start with the 15kb bin size, immediately following the QDNAseq processing I check the variance of the raw data.  If it's greater than the median of the training data (`r median(training.raw$sd)`) then I use the next greater bin size for that sample.  I do this up to a bin size of 100kb currently.

```{r, echo=F}
library(BarrettsProgressionRisk)

analysis.dir = '~/Data/BarrettsProgressionRisk/Analysis/validation/resegment/'

per.sample.var = do.call(bind_rows, lapply(list.files(analysis.dir, 'variance.tsv', recursive=T, full.names=T), function(f) {
  read_tsv(f, col_types='cddd') %>% mutate(patient = basename(sub('plots','',dirname(f))))
})) %>% dplyr::select(patient, Samples, everything())

max.bin<-function(bin15,bin50,bin100) {
  if (!is.na(bin100))
    return(100)
  if (!is.na(bin50))
    return(50)
  return(15)
}

per.sample.bin = per.sample.var %>% group_by(Samples) %>% dplyr::summarise( max.bin = max.bin(`15kb`,`50kb`,`100kb`) ) %>% arrange(max.bin)

per.sample.bin %>% group_by(max.bin) %>% tally() %>% spread(max.bin,n) %>% kable(caption = 'Number of samples per QDNAseq bin size') %>% kable_styling('striped', full_width = F)
```

@@

per.sample.bin = val.info %>% dplyr::select(matches('Hospital|Status|Block|Pathology|Samplename')) %>% 
  left_join(per.sample.bin, by=c('Samplename'='Samples'))

per.sample.bin %>% group_by(max.bin, Status) %>% tally()

per.sample.bin %>% filter(is.na(max.bin))

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(reshape2)
library(gridExtra)
library(kableExtra)
source('lib/load_patient_metadata.R')

info.file = '~/Data/BarrettsProgressionRisk/QDNAseq/training/All_patient_info.xlsx'
train.info = read.patient.info(info.file)$info

qdna.dirs = '~/Data/BarrettsProgressionRisk/Analysis/pcf_perPatient/'
train.dists.by.kb = purrr::map( list.files(qdna.dirs, full.names=T), function(dir) {
  do.call(bind_rows, lapply(list.files(dir, 'raw_dist.tsv', full.names = T, recursive = T), function(f) {
    read_tsv(f,col_types = 'ccdddddddd') %>% mutate(kb = basename(dir))  
  }))
})
names(train.dists.by.kb) = basename(list.files(qdna.dirs, full.names=T))

train.resids = purrr::map( list.files(qdna.dirs, full.names=T), function(dir) {
  do.call(bind_rows, lapply(list.files(dir, 'residuals.tsv', full.names = T, recursive = T), function(f) {
    read_tsv(f,col_types = 'ccddddl') %>% mutate(kb = basename(dir))  
  }))
})
names(train.resids) = basename(list.files(qdna.dirs, full.names=T))

@@

qdna.dirs = '~/Data/BarrettsProgressionRisk/Analysis/validation/pcf_perPatient'
val.train.dists.by.kb = purrr::map( list.files(qdna.dirs, full.names=T), function(dir) {
  do.call(bind_rows, lapply(list.files(dir, 'raw_dist.tsv', full.names = T, recursive = T), function(f) {
    read_tsv(f,col_types = 'ccdddddddd') %>% mutate(kb = basename(dir))  
  }))
})
names(val.train.dists.by.kb) = basename(list.files(qdna.dirs, full.names=T))

val.resids = purrr::map( list.files(qdna.dirs, full.names=T), function(dir) {
  do.call(bind_rows, lapply(list.files(dir, 'residuals.tsv', full.names = T, recursive = T), function(f) {
    read_tsv(f,col_types = 'ccddddl') %>% mutate(kb = basename(dir))  
  }))
})
names(val.resids) = basename(list.files(qdna.dirs, full.names=T))
```

@@

## Training Data

The training data was all reprocessed at different QDNAseq bin sizes

```{r}
m = melt(do.call(bind_rows, train.dists.by.kb), id.vars='kb', measure.vars=c('median'))
m = m %>% mutate(kb = factor(kb, levels=c('15kb','50kb','100kb','500kb'), ordered=T))
ggplot(m, aes(kb, value, group=kb)) + geom_boxplot(aes(fill=kb), show.legend = F) + ylim(0.95,1.05) + 
  scale_fill_brewer(palette='Set1') + 
  labs(title='Training data median per QDNAseq bin size', y='median',x='') + theme_minimal()

m = melt(do.call(bind_rows, train.dists.by.kb), id.vars='kb', measure.vars=c('stdev'))
m = m %>% mutate(kb = factor(kb, levels=c('15kb','50kb','100kb','500kb'), ordered=T))
ggplot(m, aes(kb, value, group=kb)) + geom_boxplot(aes(fill=kb), show.legend = F) + ylim(0,1) +
  scale_fill_brewer(palette='Set1') + 
  labs(title='Training data variance per QDNAseq bin size', y='stdev',x='') + theme_minimal()

m %>% group_by(kb) %>% dplyr::summarise( outliers = length(which(value > 1)), median.variance = median(value), min.var=min(value), max.var=max(value)) %>% mutate_if(is.numeric, list(~round(.,3))) %>%
  kable(caption='Training data variance') %>% kable_styling(full_width = F)
```

## Validation Data

The validation data was all reprocessed at different QDNAseq bin sizes

```{r}
m = melt(do.call(bind_rows, val.dists.by.kb), id.vars='kb', measure.vars=c('median'))
m = m %>% mutate(kb = factor(kb, levels=c('15kb','50kb','100kb','500kb'), ordered=T))
ggplot(m, aes(kb, value, group=kb)) + geom_boxplot(aes(fill=kb), show.legend = F) + ylim(0.95,1.05) + 
  scale_fill_brewer(palette='Set1') + 
  labs(title='Validation data median per QDNAseq bin size', y='median',x='') + theme_minimal()

m = melt(do.call(bind_rows, val.dists.by.kb), id.vars='kb', measure.vars=c('stdev'))
m = m %>% mutate(kb = factor(kb, levels=c('15kb','50kb','100kb','500kb'), ordered=T))
ggplot(m, aes(kb, value, group=kb)) + geom_boxplot(aes(fill=kb), show.legend = F) + ylim(0,1) +
  scale_fill_brewer(palette='Set1') + 
  labs(title='Validation data variance per QDNAseq bin size', y='stdev',x='') + theme_minimal()

m %>% group_by(kb) %>% dplyr::summarise( outliers = length(which(value > 1)), median.variance = median(value), min.var=min(value), max.var=max(value)) %>% mutate_if(is.numeric, list(~round(.,3))) %>%
  kable(caption='Validation data variance') %>% kable_styling(full_width = F)
```

## Training vs Validation

At every bin size there is an overall increase in the variance for the validation set.  It is not driven by the outliers.

```{r}
all = bind_rows(do.call(bind_rows, train.dists.by.kb) %>% mutate(cohort='training'),
                do.call(bind_rows, val.dists.by.kb) %>% mutate(cohort = 'validation'))
all = all %>% mutate(kb = factor(kb, levels=c('15kb','50kb','100kb','500kb'), ordered=T))

m = melt(all, id.vars=c('kb','cohort'), measure.vars='stdev')
ggplot(m, aes(cohort, value, group=cohort)) + facet_grid(~kb) + ylim(0,1) +
  geom_jitter(width=0.2,alpha=0.5) + geom_violin(aes(fill=kb), alpha=0.5, show.legend=F) + 
  #geom_boxplot(aes(fill=kb), outlier.color=NA, show.legend = F) 
  scale_fill_brewer(palette='Set1') + labs(title='Data variance per QDNAseq bin size', y='stdev',x='') + theme_minimal()

m = melt(all, id.vars=c('kb','cohort'), measure.vars='median')
ggplot(m, aes(cohort, value, group=cohort)) + facet_grid(~kb) + 
  geom_jitter(width=0.2,alpha=0.5) + geom_violin(aes(fill=kb), alpha=0.5, show.legend=F) + ylim(0.95,1.1) +
  scale_fill_brewer(palette='Set1') + labs(title='Median per QDNAseq bin size', y='median',x='') + theme_minimal()

ggplot(all %>% filter(kb == '100kb'), aes(stdev)) + facet_grid(~cohort) + 
  geom_density(fill='lightblue',kernel='gaussian') + xlim(0,1)

@@

## Post segmentation residuals

The variance continues (as before) and is reflected in the residuals of the segmented values for each bin size

@@

plist = lapply(train.resids, function(df) {
  lim = c(0, max(df$varMAD_median))
  if (max(df$varMAD_median) > 0.05) lim = c(min(df$varMAD_median), 0.05)
  ggplot(df, aes(varMAD_median)) + geom_histogram(color='lightblue') + labs(title=unique(df$kb)) + xlim(lim)
})
do.call(grid.arrange, c(plist))

all.resids = bind_rows(
  do.call(bind_rows, train.resids) %>% mutate(cohort='training'),
  do.call(bind_rows, val.resids) %>% mutate(cohort='validation') )

m = melt(all.resids, id.vars=c('kb','cohort'), measure.vars = 'varMAD_median')
ggplot(m %>% filter(value < 0.1), aes(cohort, value)) + facet_grid(~kb) + 
  geom_jitter(width=0.2, alpha=0.5) + geom_violin(aes(fill=cohort), alpha=0.5, show.legend=F) + 
  labs(title='Post-segmentation MAD residuals', x='median MAD variance')

cutoffs = list('100kb' = 0.009, '15kb' = 0.011, '500kb' = 0.005, '50kb' = 0.011)

sapply(names(train.resids), function(kb) nrow( train.resids[[kb]] %>% filter(varMAD_median >= cutoffs[kb])))
sapply(names(val.resids), function(kb) nrow( val.resids[[kb]] %>% filter(varMAD_median >= cutoffs[kb])))

ggplot(do.call(bind_rows,train.resids), aes(kb, varMAD_median)) + geom_boxplot() + ylim(0,0.05)
```
